---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi = 300
)
```

# ggbraid

<!-- badges: start -->
<!-- badges: end -->

ggbraid provides a new stat, `stat_braid()`, that extends the functionality of `geom_ribbon()` to correctly fill the area between two alternating lines (or steps) with two different colors. ggbraid also provides a geom, `geom_braid()`, that wraps `geom_ribbon()` and uses `stat_braid()` by default.

## Installation

You can install the development version of ggbraid from GitHub with:

``` r
# install.packages("devtools")
devtools::install_github("nsgrantham/ggbraid")
```

## Example

```{r example}
library(ggplot2)
library(ggbraid)
library(tidyr)

set.seed(42)  # for reproducibility

n <- 21

df_long <- tibble(
  x = c(1:n, 1:n),
  y = c(rnorm(n), rnorm(n, mean = 0.5)),
  z = c(rep("a", n), rep("b", n))
)

df_long
```

```{r pivot-wider}
df_wide <- pivot_wider(df_long, names_from = z, values_from = y)

df_wide
```

```{r geom-ribbon-without-fill}
ggplot() +
  geom_line(aes(x, y, linetype = z), data = df_long) +
  geom_ribbon(aes(x, ymin = a, ymax = b), data = df_wide, alpha = 0.2) + 
  guides(linetype = "none")
```

```{r geom-ribbon-with-fill}
ggplot() +
  geom_line(aes(x, y, linetype = z), data = df_long) +
  geom_ribbon(aes(x, ymin = a, ymax = b, fill = a < b), data = df_wide, alpha = 0.6) +
  guides(linetype = "none", fill = "none")
```

```{r geom-braid-with-fill}
ggplot() +
  geom_line(aes(x, y, linetype = z), data = df_long) +
  geom_braid(aes(x, ymin = a, ymax = b, fill = a < b), data = df_wide, alpha = 0.6) +
  guides(linetype = "none", fill = "none")
```
